---
title: "36-467 Homework 12"
author:
- Eu Jing Chua
- eujingc
date: "December 3, 2018"
output:
  pdf_document: default
header-includes:
    - \usepackage{enumerate}
    - \usepackage{longtable}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

\newcommand{\est}[1]{\hat{#1}}
\newcommand{\betah}[1]{\est{\beta_{#1}}}
\newcommand{\avg}[1]{\overline{#1}}
\newcommand{\E}[1]{\mathbb{E} \left[ #1 \right]}
\newcommand{\Var}[1]{\text{Var} \left[ #1 \right]}
\newcommand{\Cov}[1]{\text{Cov} \left[ #1 \right]}
\newcommand{\X}{\mathbb{X}}
\newcommand{\sumTo}[1]{\sum^{#1}_{i=1}}
\newcommand{\sumjTo}[1]{\sum^{#1}_{j=1}}
\newcommand{\matr}[1]{\mathbf{#1}}

## Question 1

```{r}
library(knitr)
library(kableExtra)
```


```{r cache = TRUE}
dicty <- read.table("http://www.stat.cmu.edu/~cshalizi/dst/18/hw/12/dicty-seq-1.dat")
table.dicty <- table(dicty)
kable(table.dicty,
      caption = "Counts of each base",
      longtable = TRUE)
```

The sequence has `r nrow(dicty)` bases.

## Question 2

```{r cache = TRUE}
# From Homework 12 handout
symbseq.to.blocks <- function(s, L) {
    n <- length(s)
    collapser <- function(i) {
        paste(s[i:(i + L - 1)], collapse = "")
    }
    max.index <- n - L + 1
    blocks <- sapply(1:max.index, collapser)
    return(blocks)
}

succ.pairs <- symbseq.to.blocks(dicty$V1, 2)
succ.pairs.table <- table(succ.pairs)
```
```{r}
kable(succ.pairs.table,
      caption = "Counts of all successive pairs of bases",
      longtable = TRUE)
```

## Question 3

```{r}
bases = c("A", "C", "G", "N", "T")
counts <- matrix(succ.pairs.table, nrow = 5, ncol = 5, byrow = TRUE)
colnames(counts) <- bases
rownames(counts) <- bases
q <- matrix(nrow = 5, ncol = 5)
colnames(q) <- bases
rownames(q) <- bases

for (i in bases) {
    total <- sum(counts[i, ])
    q[i, ] <- counts[i, ] / total
}
kable(q, digits = 5,
      caption = "Transition Matrix of First-Order Markov Chain",
      longtable = TRUE)
```

## Question 4

```{r}
p.init <- table.dicty / sum(table.dicty)
n <- nrow(dicty)
log.likelihood.1 <- log(p.init[dicty$V1[1]]) + sum(counts * log(q))
```

The log-likelihood of the estimated first-order Markov chain is $`r signif(log.likelihood.1, 5)`$.

## Question 5

```{r}
q.upper <- matrix(nrow = 5, ncol = 5)
colnames(q.upper) <- bases
rownames(q.upper) <- bases
q.lower <- matrix(nrow = 5, ncol = 5)
colnames(q.lower) <- bases
rownames(q.lower) <- bases
for (i in bases) {
    q.upper[i, ] <- qbinom(p = 0.975, size = table.dicty[i], prob = q[i, ])
    q.upper[i, ] <- q.upper[i, ] / sum(counts[i, ])
    q.lower[i, ] <- qbinom(p = 0.025, size = table.dicty[i], prob = q[i, ])
    q.lower[i, ] <- q.lower[i, ] / sum(counts[i, ])
}

```
The upper and lower bounds of 95% C.I. for each entry in the transition matrix is as follows:
```{r}
kable(q.upper, "markdown", digits = 5,
      caption = "Upper bound of 95% C.I. of transition matrix")

kable(q.lower, "markdown", digits = 5,
      caption = "Lower bound of 95% C.I. of transition matrix")
```

These intervals seem unusually small, being very tight.


## Question 6

```{r}
q.L.eigen <- eigen(t(q))
results <- matrix(c(Re(q.L.eigen$vectors[, 1] / sum(q.L.eigen$vectors[, 1])),
                    p.init),
                  ncol = 2, nrow = 5)
colnames(results) <- c("Invariant Dist.", "Overall Dist.")
rownames(results) <- bases
kable(results, digits = 5,
      caption = "Distributions of the Markov chain",
      longtable = TRUE)
```

From the above table, we can see that the distributions are approximately the same.

\newpage

## Question 7

```{r cache = TRUE}
succ.triples <- symbseq.to.blocks(dicty$V1, 3)
succ.triples.table <- table(succ.triples)

bases.pairs <- c(t(outer(bases, bases, paste, sep = "")))
q <- matrix(nrow = 25, ncol = 25)
q[, ] <- 0
colnames(q) <- bases.pairs
rownames(q) <- bases.pairs

for (ij in bases.pairs) {
    total <- sum(succ.triples.table[paste(ij, bases, sep = "")], na.rm = TRUE)
    for (k in bases) {
        jk <- paste(substr(ij, 2, 2), k, sep = "")
        ijk <- paste(ij, k, sep = "")
        if (ijk %in% names(succ.triples.table)) {
            q[ij, jk] <- succ.triples.table[ijk] / total
        }
    }
}
```
```{r}
kable(q, "latex",
      caption = "Transition Matrix of Second-Order Markov Chain",
      digits = 2) %>%
    kable_styling(latex_options = c("scale_down", "hold_position"))
    
```

## Question 8

```{r}
counts <- matrix(nrow = 25, ncol = 25)
counts[, ] <- 0
colnames(counts) <- bases.pairs
rownames(counts) <- bases.pairs
for (triple in names(succ.triples.table)) {
    ij <- substr(triple, 1, 2)
    jk <- substr(triple, 2, 3)
    counts[ij, jk] <- succ.triples.table[triple]
}

# Can use na.rm as we only want to sum over valid combinations of ijk only
log.likelihood.2 <- log(p.init[dicty$V1[1]]) + sum(counts * log(q), na.rm = TRUE)
```

The log-likelihood of the estimated second-order Markov chain is $`r signif(log.likelihood.2, 5)`$.

## Question 9

```{r}
test.stat <- 2 * (log.likelihood.2 - log.likelihood.1)
dof <- 5 * 5 * 4 - 5 * 4
p.value <- pchisq(test.stat, df = dof, lower.tail = FALSE)
```

We test:
\begin{align}
H_0: \text{The model is a first-order Markov chain} \\
H_a: \text{The model is a second-order Markov chain}
\end{align}

Our test statistic is $`r signif(test.stat, 5)`$, which follows a $\chi^2$ distribution with degrees of freedom `r dof`. This gives us a p-value very close to $`r signif(p.value, 5)`$, hence we reject $H_0$ and conclude that there is sufficient evidence that the model is actually a second-order Markov chain.

## Question 10

```{r cache = TRUE}
dicty.2 <- read.table("http://www.stat.cmu.edu/~cshalizi/dst/18/hw/12/dicty-seq-2.dat")

succ.pairs.2 <- symbseq.to.blocks(dicty.2$V1, 2)
succ.pairs.table.2 <- table(succ.pairs.2)
counts <- matrix(succ.pairs.table.2, nrow = 5, ncol = 5, byrow = TRUE)
colnames(counts) <- bases
rownames(counts) <- bases
q <- matrix(nrow = 5, ncol = 5)
colnames(q) <- bases
rownames(q) <- bases

for (i in bases) {
    total <- sum(counts[i, ])
    q[i, ] <- counts[i, ] / total
}

q.upper <- matrix(nrow = 5, ncol = 5)
colnames(q.upper) <- bases
rownames(q.upper) <- bases
q.lower <- matrix(nrow = 5, ncol = 5)
colnames(q.lower) <- bases
rownames(q.lower) <- bases
for (i in bases) {
    q.upper[i, ] <- qbinom(p = 0.975, size = table.dicty[i], prob = q[i, ])
    q.upper[i, ] <- q.upper[i, ] / sum(counts[i, ])
    q.lower[i, ] <- qbinom(p = 0.025, size = table.dicty[i], prob = q[i, ])
    q.lower[i, ] <- q.lower[i, ] / sum(counts[i, ])
}
```
```{r}
kable(q, digits = 5,
      caption = "Transition Matrix of First-Order Markov Chain for Second set of Chromosomes",
      longtable = TRUE)
```
The upper and lower bound of the 95% C.I. for each entry in the transition matrix is as follows:
```{r}
kable(q.upper, "markdown", digits = 5,
      caption = "Upper bound of 95% C.I. of transition matrix")

kable(q.lower, "markdown", digits = 5,
      caption = "Lower bound of 95% C.I. of transition matrix")
```


## Question 11
```{r}
table.dicty.2 <- table(dicty.2)
p.init.2 <- table.dicty.2 / sum(table.dicty.2)
n.2 <- nrow(dicty.2)
log.likelihood.3 <- log(p.init.2[dicty.2$V1[1]]) + sum(counts * log(q))
log.likelihood.sum <- log.likelihood.1 + log.likelihood.3
```

The log-likelihood of the first-order Markvoc chain for the second set of chromosomes is $`r signif(log.likelihood.3, 5)`$. The total log-likelihood of the first-order Markov chains of both sets of chromosomes is $`r signif(log.likelihood.sum, 5)`$.

## Question 12
```{r cache = TRUE}
counts <- matrix(succ.pairs.table + succ.pairs.table.2, nrow = 5, ncol = 5, byrow = TRUE)
colnames(counts) <- bases
rownames(counts) <- bases
q <- matrix(nrow = 5, ncol = 5)
colnames(q) <- bases
rownames(q) <- bases

for (i in bases) {
    total <- sum(counts[i, ])
    q[i, ] <- counts[i, ] / total
}
```
```{r}
kable(q,
      caption = "Transition Matrix of First-Order Markov Chain for both chromosomes",
      longtable = TRUE)
```

## Question 13

```{r}
p.init.pooled <- (table.dicty + table.dicty.2) / sum(table.dicty + table.dicty.2)
n.2 <- nrow(dicty) + nrow(dicty.2)
# Since both start with "N", we can use that as the initial
log.likelihood.3 <- log(p.init.pooled[dicty$V1[1]]) + sum(counts * log(q))
```

The log-likelihood for the pooled fisrt-order Markov chain is $`r signif(log.likelihood.3, 5)`$.

## Question 14

```{r}
test.stat <- 2 * (log.likelihood.sum - log.likelihood.3)
dof <- 2 * 5 * 4 - 5 * 4
p.value <- pchisq(test.stat, df = dof, lower.tail = FALSE)
```

We test:
\begin{align}
H_0: \text{Both sets of chromosomes are modelled with a shared first-order Markov chain} \\
H_a: \text{Both sets of chromosomes are individually modelled with first-order Markov chains}
\end{align}

Our test statistic is $`r signif(test.stat, 5)`$, which follows a $\chi^2$ distribution with degrees of freedom `r dof`. This gives us a p-value of $`r signif(p.value, 5)`$ which is close to 0, hence we reject $H_0$ at 0.05 significance and conclude that there is sufficient evidence that both sets of chromosomes can actually be modelled with a shared first-order Markov chain.
