---
title: "36-467 Homework 11"
author:
- Eu Jing Chua
- eujingc
date: "November 19, 2018"
output:
  pdf_document: default
header-includes:
    - \usepackage{enumerate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

\newcommand{\est}[1]{\hat{#1}}
\newcommand{\betah}[1]{\est{\beta_{#1}}}
\newcommand{\avg}[1]{\overline{#1}}
\newcommand{\E}[1]{\mathbb{E} \left[ #1 \right]}
\newcommand{\Var}[1]{\text{Var} \left[ #1 \right]}
\newcommand{\Cov}[1]{\text{Cov} \left[ #1 \right]}
\newcommand{\X}{\mathbb{X}}
\newcommand{\sumTo}[1]{\sum^{#1}_{i=1}}
\newcommand{\sumjTo}[1]{\sum^{#1}_{j=1}}
\newcommand{\matr}[1]{\mathbf{#1}}

```{r}
data <- read.csv("http://www.stat.cmu.edu/~cshalizi/statcomp/14/labs/07/ckm_nodes.csv")
```

## Q1

```{r}
library(knitr)

# Remove entries with adoption_date == NA
data <- subset(data, !is.na(adoption_date))
kable(nrow(data), caption = "Number of doctors without unknown adoption dates")
```

## Q2
```{r}
adoption.start.counts <- table(data$adoption_date)[1:17]
plot(1:17, adoption.start.counts,
     xlab = "Month", ylab = "Count",
     main = "Number of doctors who started prescribing in each month", type = "o")
```

## Q3
```{r}
adoption.start.cum <- cumsum(adoption.start.counts)
plot(1:17, adoption.start.cum,
     xlab = "Month", ylab = "Cumulative Count",
     main = "Cumulative number of doctors who prescribed by end of each month", type = "o")
```

## Q4
```{r}
adoption.potential.counts <- vector(length = 17)
for (i in 1:17) {
    adoption.potential.counts[i] <- sum(data$adoption_date > i)
}
plot(1:17, adoption.potential.counts,
     xlab = "Month", ylab = "Count", col = "red",
     ylim = c(min(adoption.potential.counts), nrow(data)),
     main = "Number of doctors who did not prescribe by end of each month", type = "o")
lines(1:17, adoption.potential.counts + adoption.start.cum, type = "o",
      col = "blue")
legend("right", legend = c("Count of non-prescribing doctors", "Sum with previous plot"),
       col = c("Red", "Blue"),
       lty = 1, pch = 1)
```

## Q5
```{r}
adoption.fraction <- adoption.start.counts[2:17] / adoption.potential.counts[1:16]
plot(2:17, adoption.fraction,
     yaxt = "n",
     xlab = "Month", ylab = "Fraction",
     main = "Fraction of doctors who started prescribing each month", type = "o")
axis(2, at = adoption.fraction, labels = round(adoption.fraction, 2))
```

## Q6
```{r}
plot(adoption.start.cum[1:16], adoption.fraction,
     yaxt = "n",
     xlab = "Cumulative count of already prescribing doctors",
     ylab = "Fraction of doctors who started prescribing", type = "o")
axis(2, at = adoption.fraction, labels = round(adoption.fraction, 2))
```

The shape is shows an increasing fraction of doctors who start prescribing as more doctors already prescribe. However, this is only up to a certain point, where it sharply decreases, then starts roughly increasing again, followed finally by another sharp decrease. This resembles some form of a damped oscillation.

## Q7
```{r}
library(latex2exp)
plot(adoption.start.cum[1:16], adoption.fraction,
     yaxt = "n",
     xlab = TeX("X_I(t)"),
     ylab = TeX("\\hat{p}_{SI}(t+1)"), type = "o",
     main = TeX("Plot of \\hat{p}_{SI}(t+1) against X_I(t)"))
axis(2, at = adoption.fraction, labels = round(adoption.fraction, 2))
```

$\est{p}_{SI}(t+1)$ is represented by `adoption.fraction`, as it counts the proportion of those who had not been prescribing before, but transitioned into prescribing for each month. $X_I(t)$ is represented by `adoption.start.cum` in the code, as it counts the total number of doctors prescribing by the end of each month.

## Q8
```{r}
p.SI <- adoption.fraction
p.SI.var <-  p.SI * (1 - p.SI) / adoption.potential.counts[1:16]
plot(2:17, p.SI.var, type = "o",
     xlab = "Month", ylab = TeX("\\hat{Var}(\\hat{p}_{SI})"),
     main = TeX("\\hat{Var}(\\hat{p}_{SI}) for each month"))
kable(max(p.SI.var) / min(p.SI.var),
      caption = "Ratio of largest to smallest estimated variance")
```

## Q9
```{r}
X.I <- adoption.start.cum
lm.fit <- lm(p.SI ~ 0 + X.I[1:16])
alpha <- coef(lm.fit)[[1]]

kable(alpha,
      caption = "Estimate of $\\alpha$")
```

\newpage

## Q10
```{r echo=TRUE}
sim.adoption <- function(alpha, n, X.I.init, months) {
    # alpha: weight such that p_SI(t) = alpha * X_I(t-1)
    # n: total number in population
    # X.I.init: Initial number of "infected"
    # months: number of months to simulate, where each time step is a month
    # Returns: Vector of counts of total infected by the month as indexed

    X.I <- vector(length = months)

    # Initial counts of "Infected"
    X.I[1] <- X.I.init

    # Edge case
    if (months < 2) {
        return(X.I)
    }

    # Simulate rest of the values
    for (t in 2:months) {
        # Count of "Susceptible" at t-1
        X.S <- n - X.I[t - 1]

        # Count of those transitioning from S to I follows binomial dist.
        # with m = X.S(t-1) and p.SI(t) = alpha * X.I(t-1)
        changed <- rbinom(1, X.S, alpha * X.I[t - 1])

        # Update those who changed from S to I
        X.I[t] <- X.I[t - 1] + changed
    }

    return(X.I)
}
```

## Q11
```{r fig.height = 4}
sims.X.I <- replicate(100, sim.adoption(alpha, nrow(data), adoption.start.counts[[1]], 17))
sim.percentiles <- apply(sims.X.I, 1, quantile, probs = c(0.05, 0.95))

matplot(1:17, t(sim.percentiles),
        ylim = c(min(sim.percentiles, adoption.start.cum),
                 max(sim.percentiles, adoption.start.cum)),
        type = "o", pch = 1, col = c("blue", "red"), lty = 2,
        xlab = "Month", ylab = "Cumulative Count",
        main = "Cumulative count of doctors who precribed by each month")
lines(1:17, adoption.start.cum, type = "o")
legend("topleft",
       legend = c("95% Percentile of Simulations",
                  "5% Percentile of Simulations",
                  "Actual Data"),
       col = c("red", "blue", "black"),
       lty = c(2, 2, 1), pch = 1, cex = 0.7)
```

The plot shows that the model, using the estimated value of $\alpha = `r signif(alpha, 3)`$, does not work well as the actual data is consistently above the $5^{th}$ and $95^{th}$ percentile of the simulations, hence is consistently underestimating.

## Q12
```{r fig.height = 4}
sims.X.I <- replicate(100, sim.adoption(2 * alpha, nrow(data), adoption.start.counts[[1]], 17))
sim.percentiles <- apply(sims.X.I, 1, quantile, probs = c(0.05, 0.95))

matplot(1:17, t(sim.percentiles),
        ylim = c(min(sim.percentiles, adoption.start.cum),
                 max(sim.percentiles, adoption.start.cum)),
        type = "o", pch = 1, col = c("blue", "red"), lty = 2,
        xlab = "Month", ylab = "Cumulative Count",
        main = "Cumulative count of doctors who precribed by each month")
lines(1:17, adoption.start.cum)
lines(1:17, adoption.start.cum, type = "o")
legend("topleft",
       legend = c("95% Percentile of Simulations",
                  "5% Percentile of Simulations",
                  "Actual Data"),
       col = c("red", "blue", "black"),
       lty = c(2, 2, 1), pch = 1, cex = 0.7)
```

Since our model with $\alpha = `r signif(alpha, 3)`$ consistently underestimates compared to the actual data, increasing $\alpha$ would cause the model to estimate more transitions from non-prescription to prescriptions with a higher probability, counteracting the effect of the underestimation. Hence, this model using $2 * \alpha$ seems to perform better, where more of the actual data seems to lie close to or within the bounds of the simulations.


```{r}
sim.model <- function(a,n,x1,t){
    sim.vector <- rep(0,n)
    sim.vector[1:x1] = 1
    for (i in 2:t){
        xt = length(which(sim.vector!=0))
        pt = a*xt
        remain = length(which(sim.vector==0))
        new.add = rbinom(prob=pt,size=remain,n=1)
        sim.vector[which(sim.vector==0)[1:new.add]] = i
    }
    num.doctors <- rep(0,(t+1))
    #Doctors never adopted
    num.doctors[t+1] = length(which(sim.vector==0))
    for (j in 1:t){
        num.doctors[j] = length(which(sim.vector<=j & sim.vector >0))
    }
    return (num.doctors)
}
```

